---
title: "Cox proportional hazards"
author: "Nathan Constantine-Cooke"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  github_document: default
  html_document:
    code_folding: hide
    toc: yes
    toc_float:
      collapsed: no
    number_sections: no
    css: ../style.css
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

This document demonstrates how to construct a Cox proportional hazards model,
assess the output, test for violations of the proportional hazards assumption,
and how to assess discriminative ability/ predictive accuracy

We will be using the `survival` package to construct the Cox model and for its
included `ovarian` data-set. We will be using the `survminer` package for 
visualisation purposes.

```{r}
library(survival, quietly = TRUE )
library(survminer, quietly = TRUE)
library(dplyr, warn.conflicts = FALSE)
```

The `ovarian` data-set has been produced from a randomised trial which compared
two treatments for ovarian cancer. `futime` denotes the survival/ censoring
time, `fustat` indicates censoring status, `rx` indicates treatment group.

```{r}
data("ovarian")
knitr::kable(head(ovarian))
```

## Constructing a Cox proportional hazards model

We will dichotomise age. We also acknowledge that we expect older patients to
have worse outcomes, so will set young age as the first level, (I.E the
reference). Inspecting a histogram of age suggests age $\geq$ 50 is a natural
cutoff. 

```{r}
hist(ovarian$age)
```

```{r}
ovarian <- ovarian %>% dplyr::mutate(age_group = ifelse(age < 50, "young",
                                                        "old"))
ovarian$age_group <- factor(ovarian$age_group, levels = c("young", "old"))
```

Note that covariates do not need to be discrete but it is easier to assess PH
assumption if they are.  

Similarly, we will reorder the levels of factors to ensure HR > 1 

```{r}
ovarian$rx <- factor(ovarian$rx, levels = c(2, 1), labels = c("A", "B"))

ovarian$resid.ds <- factor(ovarian$resid.ds,  levels = c("1", "2"), 
                           labels = c("no", "yes"))

ovarian$ecog.ps <- factor(ovarian$ecog.ps,levels = c("1", "2"), 
                          labels = c("good", "bad"))
```

We create a Cox proportional hazards object in `survival` by first creating a
`Surv` object and passing this to the `coxph` function as part of a formula.

```{r}
fit.coxph <- coxph(Surv(time = futime, event = fustat) ~
                     rx + resid.ds + age_group + ecog.ps, 
                   data = ovarian)
```

## Assessing output

The summary of the Cox object gives us many of the parameters we discussed 
in the slides. The `coef` column gives us the estimates for each $\beta_i$s.
The exponential of this coefficient will give us the HR for the exposure,
adjusted for the other parameters. The Z column gives us Wald test statistics
given by $z = \frac{\beta_i}{se(\beta_i)}$. 

```{r}
summary(fit.coxph)
```

We are also given 95% confidence intervals for the HRs. A basic sanity check
shows these confidence intervals are generated as previously presented.

```{r}
cat("Lower CI for Beta_1 is", exp(1.3814  - 1.96 * 0.6448))
```

Finally we have some statistics for the overall model. Concordance is a measure
of performance and can be interpreted as the fraction of all pairs of subjects
whose predicted survival times are correctly ordered among all subjects that can
be ordered. This is equivalent to AUC.

It is very common to visualise hazard ratios and their confidence intervals
using forest plots. This can easily be generated using `survminer::ggforest()`.

```{r}
ggforest(fit.coxph, data = ovarian)
```

## Testing the proportional hazards assumption

The proportional hazards assumption can be tested via Schoenfeld residuals.

A Schoenfeld residual is defined by an observed value of $x_i$ for a subject 
minus a weighted average of the $X_i$s for the other subjects still at
risk at time $t$ where $t$ is the time of event. The weights are each subject’s
hazard. A Wald test statistic is found which results in a chi-square
distribution with one degree of freedom. 

```{r}
ph.assumption <- cox.zph(fit.coxph)
print(ph.assumption)
```

You can also plot the residuals. If the residuals follow a pattern with time,
then this is evidence for the proportional hazards assumption being violated.

```{r}
par(mfrow = c(2, 2)) 
plot(ph.assumption)
```

## Survival curves

We may also wish to draw survival curves, stratified by a covariate, such as 
`Rx` adjusted for the other variables. In this case we need to first construct a
dataframe with as many rows as we want curves. Each column will represent the
status of the covariate. For the covariate of interest, we will vary the levels.
All other covariates should use the reference level if a factor or the mean
value if continuous. 

```{R}
rx.df <- with(ovarian,
               data.frame(rx = c("A", "B"),
                          age_group = c("young", "young"),
                          resid.ds = c("no", "no"),
                          ecog.ps = c ("good", "good")
                          )
               )

```

We can then plot the curves using `ggsurvplot()`. As can be seen in the plot,
there is a high degree of overlap for the confidence intervals which can be
attributed to the low sample size.  

```{R}
fit <- survfit(fit.coxph, newdata = rx.df)
ggsurvplot(fit, conf.int = TRUE, data = "rx.df",
           legend.labs=c("Rx A", "Rx B"),
           ggtheme = theme_minimal())
```