---
title: "Kaplan-Meier"
author: "Nathan Constantine-Cooke"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  github_document: default
  html_document:
    code_folding: hide
    toc: yes
    toc_float:
      collapsed: no
    number_sections: no
    css: ../style.css
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

This document demonstrates how to use the Kaplan-Meier method with R.

We will be using the `survival` package to calculate the Kaplan-Meier statistic
and for its included `ovarian` data-set. We will be using the `survminer`
package for  visualisation purposes.

```{r}
library(survival, quietly = TRUE )
library(survminer, quietly = TRUE)
library(dplyr, warn.conflicts = FALSE)
```

The `ovarian` data-set has been produced from a randomised trial which compared
two treatments for ovarian cancer. `futime` denotes the survival/ censoring
time, `fustat` indicates censoring status, `rx` indicates treatment group.

```{r}
data("ovarian")
knitr::kable(head(ovarian))
```

## Kaplan-Meier Survival Curve

We create a Cox proportional hazards object in `survival` by first creating a
`Surv()` object which is where we define which column indicates survival/
censoring time, and which column indicates censoring status. 

```{R}
surv_object <- Surv(time = ovarian$futime, event = ovarian$fustat)
```

We then use `survfit()` to fit a Kaplan-Meier curve. When we do not
wish to stratify, we specify `surv_object ~ 1` where `surv_object `is the object
created by the `Surv()` function. Finally we use the `ggsurvplot` function to
plot our curve.

```{R}
fit1 <- survfit(surv_object ~ 1, data = ovarian)
ggsurvplot(fit1, data = ovarian)
```

## Stratified Kaplan-Meier

When we wish to perform plot Kaplan-Meier curves stratified by groups, we simply
use `~ GROUP_NAME` instead of ` ~ 1` for `survfit` where `GROUP_NAME` is the
name of the  factor we are stratifying by. In the below example, we stratify by
the drug given to the patients. By setting `pval = TRUE`, we are given the
p-value from performing a log-rank test. I also recommend setting
`conf.int = TRUE` to plot the confidence intervals for each curve.

```{R} 
fit2 <- survfit(surv_object ~ rx, data = ovarian)
ggsurvplot(fit2, data = ovarian, pval = TRUE, conf.int = TRUE)
```